% Chapter — Built-in reference GNC components (FR-25). Follows the
% mandatory FR-29 six-section template: purpose; assumptions and domain of
% validity; derivation; discretization and implementation; validation
% evidence; references. Written ahead of the implementation (Phase 6):
% this chapter is the binding specification for the built-in
% IGncComponent implementations and the plugin-boundary semantics they
% share. The implementing change adds the chapter-manifest entry and the
% \testid{} evidence table. The equation labels eq:gnc:deltaq, eq:gnc:pd,
% eq:gnc:sat, eq:gnc:werr, eq:gnc:exactrot, eq:gnc:drprop, and
% eq:gnc:cmdrate are intended to be echoed verbatim in the module source
% (FR-29 traceability).
\chapter{Built-in Reference GNC Components}
\label{ch:gnc-builtin}

\section{Purpose}
\label{sec:gnc:purpose}

This chapter specifies the built-in C++ reference components behind the
FR-25 plugin interface: (a) a quaternion-error PD attitude controller,
(b) a dead-reckoning attitude navigator driven by the
Chapter~\ref{ch:sensors-imu} gyro increments, and (c) the ascent
pitch-program guidance --- the Phase~4 open-loop pitch program of
Chapter~\ref{ch:vehicle6dof} re-expressed as an \texttt{IGncComponent}
so the same ascent flies under the GNC stack (Phase~6 exit
criterion~10). It also pins the plugin-boundary semantics every
component, built-in or user-supplied, shares: zero-order-hold commands
over the major cycle (D-5), the \texttt{latency\_cycles} shift (exit
criterion~8), and the \texttt{oracle} truth-injection flag (exit
criterion~5). These components are \emph{reference} implementations:
deliberately simple, exactly specified, and reproducible in a few lines
of Python --- exit criterion~2 requires a Python reimplementation of the
PD controller to match the C++ torques to
$< \qty{e-9}{\newton\metre}$, so every branch and operation order in
Section~\ref{sec:gnc:pd} is normative.

\section{Assumptions and domain of validity}
\label{sec:gnc:domain}

Assumptions:
\begin{enumerate}
  \item \textbf{Cycle-synchronous execution.} Components run once per
    major cycle on the D-5 schedule; all inputs are the sensor samples
    of that cycle and all outputs are held (zero-order hold) across the
    next application interval. No component sees intra-cycle states.
  \item \textbf{No truth in \texttt{GncInput}.} Components consume only
    sensor outputs, their own state, and configuration. The
    scenario-level \texttt{oracle: true} debug flag injects truth into
    \texttt{GncInput} \emph{and} stamps the log header, so an oracle
    run is identifiable from its header alone (exit criterion~5); the
    reference components themselves never read the oracle fields.
  \item \textbf{Diagonal gains, per-axis saturation.} The PD controller
    uses positive per-axis gain vectors and clamps each torque axis
    independently (equation~\eqref{eq:gnc:sat}); a saturated command is
    direction-distorted, which is the standard, simplest behavior and
    is stated rather than hidden.
  \item \textbf{Dead reckoning is uncompensated.} The dead-reckoning
    navigator applies no bias compensation and no coning correction;
    its drift is exactly the integrated gyro error plus the bounded
    single-sample non-commutativity residual of
    equation~\eqref{eq:gnc:coningbound}. It exists as the minimal
    reference navigator and as the mechanization core the
    Chapter~\ref{ch:ekf} filter shares.
  \item \textbf{Guidance restates Phase 4 math.} The ascent guidance
    reuses the pitch-program geometry of Chapter~\ref{ch:vehicle6dof}
    (equations~\eqref{eq:vehicle6dof:pitchaxis}
    and~\eqref{eq:vehicle6dof:attitude}) unchanged; this chapter adds
    only its packaging as a component and the commanded-rate
    convention.
\end{enumerate}

Domain of validity: attitude-regulation regimes with per-cycle error
rotations far from the $\pm\pi$ antipode (the sign branch of
equation~\eqref{eq:gnc:pd} handles the double cover; exactly at
$\delta q_w = 0$, $180^{\circ}$ error, the branch picks $s = +1$
deterministically and the controller still converges, though the
commanded direction at that single point is convention, not physics).
Gains are the user's stability responsibility: the PD law is globally
stabilizing for a rigid body with pure torque actuation and positive
gains (Wie~\cite{wie2008}), but actuator saturation, latency, and
flexible dynamics can destabilize aggressive tunings --- the simulator
reports what the configuration does, per the project's
no-silent-correction rule (DX-2).

\section{Derivation}
\label{sec:gnc:derivation}

\subsection{Plugin-boundary semantics}
\label{sec:gnc:interface}

Let $T_c$ be the control period and $t_k = k\,T_c$ the cycle
boundaries. At cycle $k$ the scheduler samples sensors, calls each
configured component's \texttt{update}, composes the stack's final
actuator command vector $\vv{u}_k$ (torque and/or gimbal/throttle
commands, per the actuator suite of Chapter~\ref{ch:actuators}), and
applies it as a zero-order hold. With the configured latency $L =
\texttt{latency\_cycles} \ge 0$, the command \emph{applied} over
$[t_k, t_{k+1})$ is
\begin{equation}
  \vv{u}^{\mathrm{applied}}_k =
  \begin{cases}
    \vv{u}_{k-L}, & k \ge L,\\[2pt]
    \vv{u}_{\mathrm{neutral}}, & k < L,
  \end{cases}
  \label{eq:gnc:latency}
\end{equation}
where the neutral command is zero torque, zero gimbal deflection, and
throttle hold at the sequence-commanded value. The shift applies once, at
the GNC-to-plant boundary, not per component. What holds by construction is
a statement about the FIFO alone: fed one command sequence, its output is
that sequence delayed by exactly $\ell$ cycles with $\ell$ neutral cycles
prepended. That is what \testid{gnc_latency_fifo_semantics} gates, in open
loop, where the input sequence is fixed.

It does \emph{not} extend to the logged history of a run, and the
distinction is easy to lose. The loop is closed, so delaying the applied
torque changes the plant trajectory; from cycle~1 onward an $L = \ell$ run
and an $L = 0$ run see different states and therefore \emph{compute}
different commands. Only cycle~0 is comparable between the two runs, which
is precisely the comparison
\testid{test_latency_two_cycles_shifts_application} makes. Read as a claim
about \texttt{gnc.cmd}, ``the $L = \ell$ history is the $L = 0$ history
delayed by $\ell$'' is false; the criterion is verified at one cycle pair
end to end, and generally only at the FIFO. Closing the general case needs
an open-loop assertion driving the cycle with a recorded command sequence
against a fixed plant and asserting
$\vv{u}_{\mathrm{applied}}[i + \ell] = \vv{u}_{\mathrm{computed}}[i]$ for
every $i$, including the drain at run end, where the last $\ell$ computed
commands are never applied and no committed test looks. Both the computed
and the applied commands are logged, which is what makes the criterion
checkable from the log at all.

\subsection{Declared error-state layout}
\label{sec:gnc:errlayout}

The \texttt{nav.err} channel records truth minus estimate in the
estimator's own state convention. Forming it needs two things: the true
state, and a reading of what each slot of the estimator's state vector
means. Only the component knows the second, and the natural arrangement
--- hand the component the truth state and let it subtract --- is the one
this interface deliberately avoids. It would place the true state in the
hands of every estimator on every cycle whether or not the scenario
enabled the oracle, contradicting the FR-24 requirement that truth is
privileged and never visible to GNC plugins, and no signature can prevent
an implementation from retaining what it is passed.

The dependency is therefore inverted. A component \emph{declares} the
layout of its state vector; the loop reads the state vector the component
publishes and performs the differencing itself. What crosses the plugin
boundary is a description of the state vector, which carries no
information about the world.

A layout is an ordered set of blocks $\{(q_j, f_j, o_j)\}$, where $q_j$
names the true quantity the block is compared against, $f_j$ the form of
the comparison, and $o_j$ the block's first index. The blocks must tile
$[0, n)$ exactly for a state of dimension $n$: a gap would be logged as
zero, and zero in an error channel reads as ``no error'' rather than
``not known''. A component that declares no layout is written no
\texttt{nav.err} channel at all, for the same reason.

For every quantity other than attitude the comparison is elementwise,
with $\vv{x}$ the published state vector and $\vv{z}$ the corresponding
true quantity:
\begin{equation}
  e_{o+i} = z_i - x_{o+i}.
  \label{eq:gnc:errdiff}
\end{equation}

Attitude is why the form is declared rather than assumed. An attitude
error is a rotation difference, not a subtraction of quaternion
components, and both the side on which it is composed and its
parameterization are conventions the estimator owns. Writing
$\quat{q}_{\mathrm{est}}$ for the estimate read out of slots
$o \ldots o+3$ (scalar-first, D-7) and $\quat{q}_{\mathrm{true}}$ for the
true attitude, the two multiplicative forms are
\begin{align}
  \delta\quat{q}^{\mathrm{local}}
    &= \quat{q}_{\mathrm{est}}^{*} \quatmul \quat{q}_{\mathrm{true}},
  \label{eq:gnc:errlocal} \\
  \delta\quat{q}^{\mathrm{global}}
    &= \quat{q}_{\mathrm{true}} \quatmul \quat{q}_{\mathrm{est}}^{*},
  \label{eq:gnc:errglobal}
\end{align}
the first resolved in the estimated body frame and the second in the
inertial frame. Both are canonicalized to the $+w$ hemisphere, since
$\quat{q}$ and $-\quat{q}$ are the same rotation and without the
convention the four logged components could flip sign between
neighbouring epochs for no physical reason.
Equation~\eqref{eq:gnc:errlocal} is exactly the multiplicative error the
reference filter of Chapter~\ref{ch:ekf} defines
in~\eqref{eq:ekf:qerr}, which is how that filter's declared layout
reproduces its previous output unchanged.

The same $\delta\quat{q}$ admits a three-component reduction by the
small-angle relation
\begin{equation}
  \delta\vv{\theta} = 2\,\mathrm{sgn}(\delta q_w)\,\delta\vv{q}_v,
  \label{eq:gnc:errrotvec}
\end{equation}
which is the reduction the consistency evaluator applies to compare an
attitude error against the covariance it describes. It is applied
\emph{downstream}, by the evaluator, and is not a form a layout may declare.

A pair of declarable rotation-vector forms once existed and was removed,
which is recorded here rather than passed over silently because the reason
constrains what may be added later. The descriptor's load-bearing invariant
is that a block's error width equals its state width: one quantity,
\texttt{error\_block\_size}, both tiles the state vector during validation
and sizes the write into \texttt{nav.err}. That holds for every form except
the rotation-vector pair, which declared three slots while the differencing
routine read four quaternion slots at the block's offset --- so a layout
placing its attitude block last could pass validation and then read one
\texttt{double} past the end of the state buffer. The consequence was
reproduced through the shipped plugin path, where the out-of-bounds value
returned as exactly zero across byte-identical reruns and was therefore
indistinguishable from correct operation.

Removal was chosen over repair because the reduction is not a lost
capability: equation~\eqref{eq:gnc:errrotvec} is already implemented and
applied by the consistency evaluator, to exactly the $n = 16$, $m = 15$ case
the reference filter of Chapter~\ref{ch:ekf} presents. What removal does not
do --- and this is the honest consequence --- is serve an estimator whose
state carries a three-parameter attitude directly, as an MRP or Gibbs
vector. That case was never served: every attitude form reads
$\quat{q}_{\mathrm{est}}$ as four consecutive state slots, and a
three-parameter state publishes no such quaternion. Serving it requires a
route for a component to supply its estimated quaternion independently of
its state layout, which is a change to the descriptor rather than an added
enumerator. See \texttt{docs/gnc\_plugins.md} and
\texttt{docs/review/phase6\_coverage\_closure.md}.

A fourth form serves an estimator that treats the four quaternion
components as ordinary state entries, as the dead-reckoning navigator of
Section~\ref{sec:gnc:deadreckoning} does. Its error is additive, with the
true quaternion first sign-aligned to the estimate's hemisphere:
\begin{equation}
  e_{o+i} =
  \mathrm{sgn}\!\left(\quat{q}_{\mathrm{true}} \cdot
    \quat{q}_{\mathrm{est}}\right) q_{\mathrm{true},i}
  - q_{\mathrm{est},i},
  \qquad i = 0,\ldots,3,
  \label{eq:gnc:erradd}
\end{equation}
where the sign convention takes $\mathrm{sgn}(0) = +1$ so that an exactly
orthogonal pair is not flipped. Alignment is what keeps the additive
difference continuous across the double cover.

The loop differences against the same published state vector it writes to
\texttt{nav.est} on that cycle, which makes the logged error provably the
error of the logged estimate rather than of a second, separately read
state.

\subsection{Quaternion-error PD attitude controller}
\label{sec:gnc:pd}

Inputs per cycle: the commanded attitude and rate
$(\quat{q}_{i2\mathrm{cmd}},\, \vv{\omega}_{\mathrm{cmd}}^{\mathrm{CMD}})$
from the guidance component, and the estimated attitude and rate
$(\hat{\quat{q}}_{i2b},\, \hat{\vv{\omega}}^{B})$ from the navigation
component. The error quaternion is the attitude of the actual body
frame relative to the commanded frame, obtained from the composition
rule~\eqref{eq:notation:quatcomp}:
\begin{equation}
  \boxed{\;
  \delta\quat{q} = \quat{q}_{\mathrm{cmd2b}}
    = \quat{q}_{i2\mathrm{cmd}}^{*} \quatmul \hat{\quat{q}}_{i2b} ,
  \;}
  \label{eq:gnc:deltaq}
\end{equation}
so $\delta\quat{q} = [1, \vv{0}]$ at zero error, and for small errors
$\delta\vv{q}_v \approx \vv{\epsilon}/2$ with $\vv{\epsilon}$ the
rotation vector from commanded to actual body frame. The rate error
resolves the commanded rate into body axes through the error DCM,
\begin{equation}
  \vv{\omega}_{\mathrm{err}} = \hat{\vv{\omega}}^{B}
    - \dcm{\mathrm{CMD}}{B}(\delta\quat{q})\,
      \vv{\omega}_{\mathrm{cmd}}^{\mathrm{CMD}} ,
  \label{eq:gnc:werr}
\end{equation}
with the DCM from equation~\eqref{eq:notation:quat2dcm}. The commanded
torque, componentwise for $i \in \{x, y, z\}$ and with the shortest-path
sign
\begin{equation}
  s = \begin{cases} +1, & \delta q_w \ge 0,\\ -1, & \delta q_w < 0,
      \end{cases}
  \label{eq:gnc:sign}
\end{equation}
is
\begin{equation}
  \boxed{\;
  \tau_i = -\,K_{p,i}\; s\,\delta q_{v,i}
           \;-\; K_{d,i}\; \omega_{\mathrm{err},i} ,
  \;}
  \label{eq:gnc:pd}
\end{equation}
followed by the per-axis clamp
\begin{equation}
  \tau_i \leftarrow \min\bigl(\max(\tau_i,\, -\tau_{\max,i}),\,
    \tau_{\max,i}\bigr).
  \label{eq:gnc:sat}
\end{equation}
Normative arithmetic, for the exit-criterion-2 reproduction: the inputs
are used as received (no renormalization of $\delta\quat{q}$); the
Hamilton product~\eqref{eq:notation:hamiltonproduct} and conjugate are
evaluated in the project's scalar-first component order; the sign
branch is exactly $\delta q_w \ge 0$ (zero maps to $+1$); products and
clamps are elementwise with the gains $K_{p,i} > 0$
(\unit{\newton\metre}, per unit quaternion-vector component) and
$K_{d,i} > 0$ (\unit{\newton\metre\second\per\radian}). Because
$\delta\vv{q}_v \approx \vv{\epsilon}/2$, the small-error proportional
stiffness is $K_{p,i}/2$ per radian --- the factor of two lives in the
gain interpretation, not in the code. The $s$ factor selects the
shorter of the two rotations representing the same error (quaternion
double cover), which removes the classical unwinding behavior; global
asymptotic stability of the law for a rigid body with positive-definite
gains is the standard quaternion-feedback result
(Wie~\cite{wie2008}; Markley and Crassidis~\cite{markley2014}).

\subsection{Dead-reckoning attitude navigation}
\label{sec:gnc:deadreckoning}

The navigator holds a unit quaternion estimate
$\hat{\quat{q}}_{i2b}$, initialized from configuration
($\hat{\quat{q}}_0$; the reference scenarios initialize it to the true
initial attitude, which the mission file states explicitly --- there is
no implicit truth access). Each cycle it composes the measured gyro
increment $\Delta\tilde{\vv{\theta}}_k$
(equation~\eqref{eq:imu:gyro}) as an exact rotation. With the exact
exponential map
\begin{equation}
  \delta\quat{q}(\vv{\phi}) =
    \Bigl[\cos\tfrac{\phi}{2},\;
      \sin\tfrac{\phi}{2}\,\frac{\vv{\phi}}{\phi}\Bigr],
  \qquad \phi = \norm{\vv{\phi}},
  \qquad
  \delta\quat{q}(\vv{0}) = [1, \vv{0}],
  \label{eq:gnc:exactrot}
\end{equation}
the propagation and rate estimate are
\begin{equation}
  \boxed{\;
  \hat{\quat{q}}_{k} = \frac{\hat{\quat{q}}_{k-1} \quatmul
      \delta\quat{q}\bigl(\Delta\tilde{\vv{\theta}}_k\bigr)}
    {\bigl\lVert \hat{\quat{q}}_{k-1} \quatmul
      \delta\quat{q}\bigl(\Delta\tilde{\vv{\theta}}_k\bigr)
      \bigr\rVert},
  \qquad
  \hat{\vv{\omega}}_k^{B} =
    \frac{\Delta\tilde{\vv{\theta}}_k}{\Delta t_s} ,
  \;}
  \label{eq:gnc:drprop}
\end{equation}
body-side composition per equation~\eqref{eq:notation:quatcomp},
normalized every cycle. The exact-rotation choice (rather than a
small-angle quaternion) is pinned because it is branch-free above the
$\phi = 0$ special case, exactly norm-preserving before rounding, and
identical to the map the error-state filter of Chapter~\ref{ch:ekf}
uses --- one shared function. Treating the increment as a rotation
vector is exact for rotation about a fixed axis within the interval;
for axis motion within the interval the per-cycle residual is the
classical single-sample coning term, bounded by
\begin{equation}
  \norm{\vv{e}_{\mathrm{coning},k}}
    \le \tfrac{1}{12}\,
      \bigl\lVert \Delta\vv{\theta}_{k-1} \times \Delta\vv{\theta}_k
      \bigr\rVert
      + O\bigl(\norm{\Delta\vv{\theta}}^4\bigr),
  \label{eq:gnc:coningbound}
\end{equation}
the leading term of the two-sample coning correction this navigator
deliberately omits (Savage~\cite{savage1998a}; Titterton and
Weston~\cite{titterton2004}). At $\qty{100}{\hertz}$ cycling and
$\norm{\vv{\omega}} \le \qty{0.1}{\radian\per\second}$ the bound is
$\lesssim \qty{8e-8}{\radian}$ per second of flight --- far below the
gyro error budget, which is the point: the reference navigator's error
is dominated by the sensor, not the algorithm.

\subsection{Ascent pitch-program guidance component}
\label{sec:gnc:ascent}

The component is configured with the same data as the Phase~4 open-loop
sequence action: flight azimuth $A$, node tables $(t_j, \theta_j)$,
$j = 1 \dots P$, and it captures the frozen launch-site tangent basis
$(\hat{\vv{u}}_0, \hat{\vv{e}}_0, \hat{\vv{n}}_0)$ at initialization
(Chapter~\ref{ch:vehicle6dof}). Per cycle at time $t$:
\begin{enumerate}
  \item \textbf{Pitch interpolation, clamped:}
    \begin{equation}
      \theta(t) =
      \begin{cases}
        \theta_1, & t \le t_1,\\
        \theta_P, & t \ge t_P,\\
        \theta_j + \dfrac{t - t_j}{t_{j+1} - t_j}
          \bigl(\theta_{j+1} - \theta_j\bigr),
          & t_j < t \le t_{j+1} \text{ (first such } j\text{)},
      \end{cases}
      \label{eq:gnc:interp}
    \end{equation}
    the exact rule of the Phase~4 implementation, restated here so both
    paths interpolate identically.
  \item \textbf{Commanded attitude:} thrust axis
    $\hat{\vv{d}}^{I}(\theta(t), A)$ from
    equation~\eqref{eq:vehicle6dof:pitchaxis} in the frozen basis, and
    $\quat{q}_{i2\mathrm{cmd}}(t)$ from the Gram--Schmidt triad of
    equation~\eqref{eq:vehicle6dof:attitude} with up-reference
    $\hat{\vv{u}}_0$, continued through the vertical by
    equation~\eqref{eq:vehicle6dof:rollref} where that reference is
    ill conditioned. The generic degeneracy fallback of
    \texttt{attitude\_from\_body\_x} is therefore unreachable from this
    path: the continuation switches two decades above the fallback's
    own guard, so the commanded roll is always resolved from the
    commanded azimuth rather than from an inertial axis. The open-loop
    mode applies the identical continuation --- the roll convention must
    match Phase~4 exactly for exit criterion~10's re-gate to compare
    like with like, and both paths moved together when it changed.
  \item \textbf{Commanded rate}, the finite-difference rotation between
    this cycle's and the next cycle's commanded attitude, resolved in
    the commanded frame at $t$:
    \begin{equation}
      \boxed{\;
      \delta\quat{q}_c = \quat{q}_{i2\mathrm{cmd}}(t)^{*} \quatmul
        \quat{q}_{i2\mathrm{cmd}}(t + T_c),
      \qquad
      \vv{\omega}_{\mathrm{cmd}}^{\mathrm{CMD}} =
        \frac{2\,\operatorname{sgn}(\delta q_{c,w})}{T_c}\,
        \delta\vv{q}_{c,v} ,
      \;}
      \label{eq:gnc:cmdrate}
    \end{equation}
    with $\operatorname{sgn}(0) = +1$: the small-angle rate extraction
    on the short branch of the double cover, exactly the Phase~4
    convention. Its $O(\norm{\delta\vv{q}_{c,v}}^3)$ small-angle error
    is negligible at per-cycle command increments
    ($\lesssim 10^{-2}$ rad).
\end{enumerate}
The component emits
$(\quat{q}_{i2\mathrm{cmd}}, \vv{\omega}_{\mathrm{cmd}}^{\mathrm{CMD}})$
for the PD controller of Section~\ref{sec:gnc:pd}; closing the loop
through the Chapter~\ref{ch:actuators} actuators is what exit
criterion~10 re-times against the FR-32 ascent performance target.

\section{Discretization and implementation}
\label{sec:gnc:implementation}

Binding implementation notes for the Phase~6 modules:
\begin{enumerate}
  \item \textbf{Modules and traceability.} Each component is a
    C++ \texttt{IGncComponent} registered in the FR-25 static registry
    and selected by configuration string; Python subclasses enter
    through the pybind11 trampoline with identical semantics. Source
    comments echo \texttt{eq:gnc:deltaq}, \texttt{eq:gnc:pd},
    \texttt{eq:gnc:sat}, \texttt{eq:gnc:werr},
    \texttt{eq:gnc:exactrot}, \texttt{eq:gnc:drprop}, and
    \texttt{eq:gnc:cmdrate} verbatim.
  \item \textbf{Statelessness inventory.} The PD controller is
    stateless; the dead reckoner's state is $\hat{\quat{q}}$; the
    guidance's state is the frozen basis and tables. All state is
    reset by \texttt{Sim.reset} (FR-24), which is what makes stepping
    and batch runs bit-identical (exit criterion~4).
  \item \textbf{No draws, no clock, no allocation.} Components consume
    no PRNG streams and never read wall-clock time; the update path
    allocates nothing (D-10).
  \item \textbf{Logged channels.} Computed and applied commands land in
    the reserved \texttt{nav.*} group; the dead reckoner logs
    $\hat{\quat{q}}$ (scalar-first) as its estimate channel. The EKF's
    richer \texttt{nav.est.*} layout is defined in
    Chapter~\ref{ch:ekf}.
  \item \textbf{Sensor-validity handling.} Components receive each
    sensor sample with its validity flag
    (Chapter~\ref{ch:sensors-optical} semantics) and must skip work
    keyed to invalid samples; the dead reckoner consumes only the IMU,
    which is always valid.
\end{enumerate}

\section{Validation evidence}
\label{sec:gnc:validation}

Tables~\ref{tab:gnc:validation}, \ref{tab:gnc:validation-crit2},
\ref{tab:gnc:validation-crit2-guards},
and~\ref{tab:gnc:validation-py} list the
tests that constitute this chapter's validation evidence. Each test
identifier is machine-checked to exist in the test tree by
\texttt{scripts/lint\_chapters.py}. The one planned gate not yet
discharged --- the exit-criterion-10 performance re-gate --- is stated
after the tables.

\begin{longtable}{lp{8.4cm}}
  \caption{Validation evidence for the built-in GNC components: core
    (doctest, \texttt{cpp/tests/test\_gnc.cpp}).}
  \label{tab:gnc:validation} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endfirsthead
  \multicolumn{2}{@{}l}{\itshape Table \thetable\ continued from the previous page.} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endhead
  \bottomrule
  \endlastfoot
    \testid{gnc_pd_attitude_golden} &
      Six committed golden cases drive
      equations~\eqref{eq:gnc:deltaq}--\eqref{eq:gnc:sat} against
      \texttt{tests/golden/gnc/pd\_attitude.toml}, whose expected torques
      are a 60-digit \texttt{mpmath} evaluation of the normative law on
      bit-identical inputs rather than a hand computation by either
      implementation. The set covers both signs of $\delta q_w$, the
      $\delta q_w = 0$ case that pins $\mathrm{sgn}(0) = +1$, and---for
      equation~\eqref{eq:gnc:sat}, which is per-axis and two-sided---cases
      that between them clamp on the upper rail, on the lower rail, and on
      all three axes, each beside an axis left strictly inside its limit.
      The doctest asserts that coverage rather than assuming it: a set
      railing one axis on one side would leave a law clamping only that axis,
      or only that sign, reproducing every expected torque. \\
    \testid{gnc_dead_reckoning_golden} &
      The dead reckoner propagates a committed increment sequence through
      equation~\eqref{eq:gnc:drprop} to its golden state; the initial
      estimate arrives as configuration, confirming the component takes no
      implicit truth access. Declared state, covariance, and innovation
      dimensions match the plugin contract. \\
    \testid{gnc_registry_resolution_and_rejection} &
      The registry exposes exactly the four built-in names; an unknown
      component raises \texttt{std::invalid\_argument} with the registered
      set in the message, and duplicate registration --- a determinism
      hazard --- raises \texttt{std::logic\_error}. \\
    \testid{gnc_attitude_hold_guidance} &
      A non-unit configured target is normalized at construction and echoed
      each update with exactly zero commanded rate; with no configured
      target the component holds the scenario initial attitude from the
      init context. \\
    \testid{gnc_pitch_program_guidance_matches_openloop_machinery} &
      The commanded attitude is computed by the same functions in the same
      call sequence as the Phase~4 open-loop pitch-program mode, against a
      reference reproducing that arithmetic verbatim on a mid-latitude pad
      basis (equation~\eqref{eq:gnc:interp}). \\
    \testid{gnc_latency_fifo_semantics} &
      Exit criterion~8 at the FIFO level: $L = 0$ is passthrough; $L = 2$
      shifts application by exactly two cycles, the pre-fill holds applying
      the neutral command with \texttt{valid == false} and zero torque
      (equation~\eqref{eq:gnc:latency}). \\
    \testid{gnc_attitude_error_width_equals_state_width_for_every_form} &
      The descriptor invariant, enumerated over every admissible attitude
      form rather than sampled: each reports four slots, which is the width
      the differencing routine reads at a block's offset. A form added with
      any other width fails here. The attitude quantity's refusal of the
      additive difference form is asserted alongside, which is what bounds
      the enumeration to the three quaternion forms. \\
    \testid{gnc_attitude_block_last_layout_cannot_outrun_the_state_buffer} &
      The construction that formerly read out of bounds, re-attempted
      through the shipped validator: a three-slot velocity block followed by
      an attitude block over a six-slot state --- the tiling a three-slot
      attitude form once satisfied --- is now rejected, and the message
      names the arithmetic (seven slots declared against six). The same
      layout at seven slots is accepted and read back, with a guard element
      past the declared width shown untouched, so the rejection is about the
      width and not a blanket refusal of an attitude-last layout. \\
    \testid{gnc_base_component_accessors_refuse_an_undeclared_state} &
      The base-class contract for a component that declares
      \texttt{state\_dim() > 0} without overriding the state or covariance
      accessor: both raise rather than let the loop log a buffer nothing
      wrote. The remaining default introspection is asserted to be well
      defined and empty, which is what makes the two throws a contract
      rather than a gap. \\
\end{longtable}

\begin{table}[htbp]
  \centering
  \caption{Validation evidence for the built-in GNC components: exit
    criterion~2, both halves and their non-vacuity guards (pytest,
    \texttt{tests/python/test\_gnc\_missions.py} and
    \texttt{test\_gnc\_pd\_golden.py}).}
  \label{tab:gnc:validation-crit2}
  \begin{tabular}{lp{7.2cm}}
    \toprule
    Test ID & Acceptance basis \\
    \midrule
    \testid{test_pd_law_python_reimplementation_contract} &
      Exit criterion~2, mission half: an independent Python transcription of
      the PD law --- importing only \texttt{numpy} and \texttt{tomllib}, and
      re-deriving the Hamilton product, the sign branch, the error DCM, and
      the clamp by hand --- reproduces the compiled built-in's logged
      applied torques across all 601 cycles of the criterion-2 scenario.
      Measured worst residual \qty{1.3878e-17}{\newton\metre} against the
      criterion's \qty{e-9}{\newton\metre}. \\
    \testid{test_python_pd_law_matches_the_mpmath_golden} &
      Exit criterion~2, golden half: the same Python transcription
      evaluated against the committed \texttt{mpmath} goldens, one
      parametrization per case, to the golden manifest's
      $\max(5\times10^{-15}\lvert\tau\rvert, 10^{-18})$ bound, with the
      criterion's own \qty{e-9}{\newton\metre} asserted alongside. Requires
      no compiled core, so this half is gated even on a core-less
      checkout. \\
    \bottomrule
  \end{tabular}
\end{table}

% longtable, not table: this grew past a single page when the
% test_golden_expectations_record_the_branches row was added, and a float
% that overfills its page silently drops the rows that do not fit.
\begin{longtable}{lp{7.2cm}}
  \caption{Validation evidence for the built-in GNC components: the
    non-vacuity guards on exit criterion~2, which assert that the two gates
    above are sensitive to every equation they claim to cover (pytest).}
  \label{tab:gnc:validation-crit2-guards} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endfirsthead
  \multicolumn{2}{@{}l}{\itshape Table \thetable\ continued from the previous page.} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endhead
  \bottomrule
  \endlastfoot
    \testid{test_golden_cases_cover_every_branch} &
      Non-vacuity of the golden half, asserted by re-evaluating the
      reference law on the committed inputs: the
      committed set contains a case with $\delta q_w < 0$, one with
      $\delta q_w$ exactly zero, one with $\delta q_w > 0$, one clamping
      some axes but not all (without which equation~\eqref{eq:gnc:sat}
      would be indistinguishable from an unsaturated law), and one
      combining a non-zero commanded rate with a non-identity error DCM
      (without which equation~\eqref{eq:gnc:werr}'s rotation would be
      untested), so a regenerated golden set that quietly lost a branch
      fails here rather than silently weakening the gate. \\
    \testid{test_golden_expectations_record_the_branches} &
      The same non-vacuity asked of the committed expectations rather
      than of the inputs, so a regeneration that kept the saturating
      inputs while recording an unclamped answer is caught too: some
      \texttt{expected\_tau\_nm} component equals its $\tau_{\max}$
      exactly while another axis of that case sits strictly inside it
      (equation~\eqref{eq:gnc:sat} assigns the limit, which is what makes
      exact binary64 equality its fingerprint and not a coincidence),
      each of the three $\delta q_w$ branches carries a non-zero recorded
      torque so none is present but inert, and every recorded quaternion
      is unit --- the premise under which $\delta q_w$ alone fixes the
      error rotation's magnitude. This file is the suite's only evidence
      for equation~\eqref{eq:gnc:sign} at $\delta q_w = 0$, a point no
      closed-loop scenario reaches. \\
    \testid{test_pd_scenario_exercises_every_equation} &
      Non-vacuity of the mission half, asserted on the measured log rather
      than assumed from the fixture's construction: commanded rate up to
      \qty{0.0133}{\radian\per\second}; error DCM reaching \num{0.71} from
      the identity and \num{1.24} from its own transpose, so neither the
      rotation nor a transposition of it is invisible; 300 of 601 cycles on
      the $\delta q_w < 0$ branch and 301 on the other; the clamp catching
      on 92 cycles; and every body axis carrying torque, so no per-axis
      gain path is multiplied by zero. \\
\end{longtable}

\begin{table}[htbp]
  \centering
  \caption{Validation evidence for the built-in GNC components: remaining
    in-loop behavior (pytest, \texttt{tests/python/}).}
  \label{tab:gnc:validation-py}
  \begin{tabular}{lp{7.2cm}}
    \toprule
    Test ID & Acceptance basis \\
    \midrule
    \testid{test_dead_reckoning_tracks_torque_driven_truth} &
      Against torque-driven truth the logged estimate error stays below
      $5\times10^{-3}$ in both the quaternion and rate components --- the
      accumulated first-order-hold bound --- and the commanded slew settles
      to the target to sub-milliradian alignment. \\
    \testid{test_pitch_program_guidance_equals_openloop_command} &
      The closed-loop ascent's commanded attitude equals the open-loop
      pitch-program attitude bit-for-bit at every shared cycle time,
      the commanded history being trajectory-independent (frozen pad basis,
      absolute-time table). \\
    \testid{test_latency_two_cycles_shifts_application} &
      Exit criterion~8 end to end: with $L = 2$ the first valid applied
      command moves from $t = 0$ to $t = 2$ cycles, the two pre-fill cycles
      apply zero torque, and the shifted command is bit-identical to the
      $L = 0$ cycle-0 command. \\
    \bottomrule
  \end{tabular}
\end{table}

\paragraph{Scope of the criterion-2 rows, and the degeneracy they were
built to escape.} The committed reference mission --- an attitude hold ---
is degenerate against three of the five equations of the PD law, and the
degeneracy is a property of the fixture rather than of the tolerance. On
that mission the commanded rate $\vv{\omega}_{\mathrm{cmd}}$ is identically
zero, so equation~\eqref{eq:gnc:werr} has nothing to rotate; the tracking
error stays well inside a half turn, so the short-path branch of
equation~\eqref{eq:gnc:sign} is never taken; and the \qty{10}{\degree}
transient never reaches $\vv{\tau}_{\max}$, so equation~\eqref{eq:gnc:sat}
never clamps. That is exact rather than merely weak: a Python reference
that dropped any of the three reproduced the logged torques identically,
so no tolerance could have separated it.

Criterion~2 therefore runs on a scenario built to escape each of those
three, driving the FR-24 external-guidance seam so the commanded attitude
and rate are the driver's to choose while the compiled built-in still
computes every torque. It carries a constant non-zero commanded rate about
one axis; an initial commanded attitude offset of \qty{60}{\degree} about a
\emph{different} axis, since parallel axes would leave
$\mat{C}(\delta\quat{q})\,\vv{\omega}_{\mathrm{cmd}} =
\vv{\omega}_{\mathrm{cmd}}$ and make both the rotation and its transpose
invisible; the commanded quaternion expressed antipodally for the first
half of the run, so that a controller honouring
equation~\eqref{eq:gnc:sign} produces the same torque from both physically
identical representations while one omitting the branch reverses it; and a
\qty{60}{\degree} offset against \qty{0.4}{\newton\metre\per\radian} gains
with \qty{0.05}{\newton\metre} of authority, so the opening transient
saturates every axis. \testid{test_pd_scenario_exercises_every_equation}
asserts each property on the measured log, so the gate cannot quietly
return to being degenerate.

The criterion's conjunction --- an independent Python controller, the
compiled controller, a golden scenario, and \qty{e-9}{\newton\metre} --- is
met by the pair of rows above rather than by either alone.
\testid{test_python_pd_law_matches_the_mpmath_golden} evaluates the Python
transcription against \texttt{tests/golden/gnc/pd\_attitude.toml}, the same
file \testid{gnc_pd_attitude_golden} drives the C++ component with and
whose expected torques come from a 60-digit \texttt{mpmath} evaluation of
the specification rather than from either implementation;
\testid{test_pd_law_python_reimplementation_contract} evaluates that same
function against the torques the compiled component logs in the loop. The
four mutations the earlier fixture could not separate are each rejected by
both rows: dropping the sign branch moves the mission-side worst residual
to \qty{0.1}{\newton\metre}, dropping the error-DCM rotation of
$\vv{\omega}_{\mathrm{cmd}}$ to \qty{4.7340e-2}{\newton\metre}, transposing
that DCM to \qty{5.1548e-2}{\newton\metre}, and dropping the clamp to
\qty{0.1526}{\newton\metre} --- each at least seven orders above the
criterion's \qty{e-9}{\newton\metre}, against an unmutated
\qty{1.3878e-17}{\newton\metre}. On the golden side the same four exceed
the manifest bound by factors between $2.3\times10^{13}$ and
$4.0\times10^{14}$.

One clause of exit criterion~10 is deferred rather than outstanding, and
the distinction matters:
\begin{itemize}
  \item \textbf{Performance re-gate (exit criterion 10), Pi~5 hardware
    clause.} The FR-32 ascent target ($\ge 100\times$ real time, single
    core) with the full built-in GNC stack in the loop is measured as the
    \texttt{ascent\_gnc\_rt\_factor} metric of
    \texttt{scripts/perf\_gate.py}, on the committed closed-loop ascent
    \texttt{missions/ascent\_leo\_gnc.toml}, and is gated on every nightly
    run because it sits in that harness's default metric set. What is
    deferred is only the requirement that the measurement be taken on real
    Pi~5 silicon, no such hardware being available. That deferral runs
    through the PRD section~9 valve to \texttt{docs/release\_checklist.md}
    item~1, procedure \texttt{docs/perf/pi5\_checklist.md} step~4, which
    measures this metric alongside the three Phase~5 absolutes in the same
    invocation. The nightly \texttt{ubuntu-24.04-arm} leg is the interim
    aarch64 proxy and is recorded as a proxy, never as a Pi~5 measurement.
    The item is registered and fully prepared --- mission, metric, and gate
    all committed and exercised in CI --- so qualifying a release against
    the clause is one scripted command once the hardware exists.
\end{itemize}

\section{References}
\label{sec:gnc:references}

The quaternion-feedback PD law and its stability properties follow
Wie~\cite{wie2008} and Markley and Crassidis~\cite{markley2014},
transcribed to the project's D-7 conventions; the strapdown increment
composition and the coning-residual bound follow
Savage~\cite{savage1998a} and Titterton and
Weston~\cite{titterton2004}; the ascent geometry is derived in
Chapter~\ref{ch:vehicle6dof}. Full bibliographic details appear in the
bibliography.
