% Chapter — Radio-class sensors: external nav fix and altimeter (FR-23,
% A-6). Follows the mandatory FR-29 six-section template: purpose;
% assumptions and domain of validity; derivation; discretization and
% implementation; validation evidence; references. Written ahead of the
% implementation (Phase 6): this chapter is the binding specification for
% the external-nav-fix and altimeter sensor modules. The implementing
% change adds the chapter-manifest entry and the \testid{} evidence table.
% The equation labels eq:radio:fix, eq:radio:gm, eq:radio:alt, and
% eq:radio:chi2 are intended to be echoed verbatim in the module source
% (FR-29 traceability).
\chapter{Sensors: External Nav Fix and Altimeter}
\label{ch:sensors-radio}

\section{Purpose}
\label{sec:radio:purpose}

This chapter specifies the two radio-class FR-23 sensors. The
\emph{external nav fix} is the project's generalization of GNSS
(assumption A-6): an abstract service that delivers position and
velocity fixes in the GCRF with configurable Gaussian errors, so the
same sensor stands in for GPS, a future LNSS-class lunar service, or
ground-based radiometric tracking without re-architecting. The
\emph{altimeter} delivers a scalar geodetic-altitude measurement over
the central body's reference ellipsoid with a per-run bias and white
noise. Both feed the Chapter~\ref{ch:ekf} reference EKF as aiding
updates, and both are gated by Phase~6 exit criterion~6: error
statistics inside two-sided 95\,\% chi-square bounds over 1{,}000 seeded
draws, with the exact statistic defined in
Section~\ref{sec:radio:stats}.

\section{Assumptions and domain of validity}
\label{sec:radio:domain}

Assumptions:
\begin{enumerate}
  \item \textbf{Fix-level abstraction.} The nav fix models the
    \emph{output} of an external navigation service, not its
    measurements: no pseudoranges, constellation geometry, dilution of
    precision, clock states, or atmospheric delays. Error anisotropy
    that a real service exhibits (e.g.\ vertical inflation) is
    expressible through the per-axis sigmas; error correlation with
    geometry is not, and analyses that depend on it are outside this
    model's domain.
  \item \textbf{GCRF axes.} Fix errors are drawn per GCRF axis. A
    service whose natural error frame is local (east/north/up) is
    approximated by choosing equivalent GCRF sigmas; the difference
    matters only for strongly anisotropic configurations.
  \item \textbf{Optional correlated component.} A first-order
    Gauss--Markov error component exists as an option on both the
    position and the velocity fix (Section~\ref{sec:radio:fix}),
    defaulted off. It models slowly varying service biases
    (ephemeris/clock residuals). When enabled, the chi-square gate of
    Section~\ref{sec:radio:stats} applies only to the white part
    (the gate scenario disables the correlated part); a separate
    deterministic test pins the recursion.
  \item \textbf{Altimeter over the reference ellipsoid.} The altimeter
    measures geodetic altitude over the central body's reference
    ellipsoid exactly as the environment models do
    (Chapter~\ref{ch:environment}): the Bowring two-pass conversion of
    Chapter~\ref{ch:harrispriester},
    equation~\eqref{eq:hp:geodetic}, applied to the Earth-fixed (or
    planet-fixed) truth position via the Chapter~\ref{ch:frames}
    rotation. Terrain, tides, and geoid undulation are not modeled ---
    the instrument is a radar altimeter over the ellipsoid, and any
    scenario needing terrain-relative altitude is out of scope.
  \item \textbf{No latency.} Fixes apply at their sample instant;
    transport and processing latency for realism studies is provided
    generically by the FR-25 \texttt{latency\_cycles} mechanism
    (Chapter~\ref{ch:gnc-builtin}), not duplicated per sensor.
\end{enumerate}

Domain of validity: any state the core can propagate; the altimeter
additionally requires the truth position inside its configured operating
band $h \in [h_{\min}, h_{\max}]$ (outside it the sample is emitted with
$\mathrm{valid} = \mathrm{false}$, the flag semantics of
Chapter~\ref{ch:sensors-optical}) and outside the Bowring conversion's
polar-axis neighborhood only in the sense documented in
Chapter~\ref{ch:harrispriester}. Non-positive sigmas, $\tau_c \le 0$, or
$h_{\min} \ge h_{\max}$ are rejected by the FR-15 validator; the in-loop
model never throws.

\section{Derivation}
\label{sec:radio:derivation}

\subsection{External nav fix}
\label{sec:radio:fix}

At each sample (major-cycle schedule, D-5), with truth position
$\vv{r}^{I}$ and velocity $\vv{v}^{I}$ (the logged \texttt{truth}
channels),
\begin{equation}
  \boxed{\;
  \vv{r}_{\mathrm{meas}} = \vv{r}^{I} + \vv{c}_{r,k} + \vv{\eta}_{r,k},
  \qquad
  \vv{v}_{\mathrm{meas}} = \vv{v}^{I} + \vv{c}_{v,k} + \vv{\eta}_{v,k},
  \;}
  \label{eq:radio:fix}
\end{equation}
with independent white errors per GCRF axis $a \in \{x,y,z\}$,
\begin{equation}
  \eta_{r,k,a} \sim \mathcal{N}(0, \sigma_{r,a}^2),
  \qquad
  \eta_{v,k,a} \sim \mathcal{N}(0, \sigma_{v,a}^2),
  \label{eq:radio:white}
\end{equation}
and the optional first-order Gauss--Markov components (defaulted off,
$\vv{c} \equiv \vv{0}$) following exactly the recursion of
Chapter~\ref{ch:sensors-imu}, equation~\eqref{eq:imu:gm}:
\begin{equation}
  \vv{c}_{k} = \varphi\,\vv{c}_{k-1} + \vv{w}_k,
  \qquad
  \varphi = e^{-\Delta t_s/\tau_c},
  \qquad
  \vv{w}_k \sim \mathcal{N}\bigl(\vv{0},\,
    \sigma_{\mathrm{GM}}^2(1 - \varphi^2)\,\mat{I}_3\bigr),
  \qquad
  \vv{c}_{0} \sim \mathcal{N}\bigl(\vv{0},\,
    \sigma_{\mathrm{GM}}^2\,\mat{I}_3\bigr),
  \label{eq:radio:gm}
\end{equation}
with separate $(\sigma_{\mathrm{GM}}, \tau_c)$ pairs for the position
and velocity components. The exactness of the discretization and its
stationary initialization are established in
Chapter~\ref{ch:sensors-imu}. The fix carries
$\mathrm{valid} = \mathrm{true}$ unconditionally in v1; outage modeling
is a configuration concern deferred until a use case demands it (PRD
Non-Goals discipline), and the flag exists in the interface so adding it
later breaks nothing.

\subsection{Altimeter}
\label{sec:radio:alt}

With $h(\vv{r})$ the geodetic altitude of the truth position over the
central body's reference ellipsoid --- the planet-fixed position
$\dcm{I}{E}\,\vv{r}^{I}$ (Chapter~\ref{ch:frames}) put through the
Bowring two-pass conversion of
equation~\eqref{eq:hp:geodetic} with the body's $(a, 1/f)$; for a
spherical body ($f = 0$) the conversion degenerates exactly to
$h = \norm{\vv{r}} - a$ --- the measurement is
\begin{equation}
  \boxed{\;
  h_{\mathrm{meas},k} = h(\vv{r}^{I}_k) + b_h + \eta_{h,k},
  \qquad
  b_h \sim \mathcal{N}(0, \sigma_{b}^2) \text{ once per run},
  \qquad
  \eta_{h,k} \sim \mathcal{N}(0, \sigma_{h}^2),
  \;}
  \label{eq:radio:alt}
\end{equation}
with validity
\begin{equation}
  \mathrm{valid}_k = \bigl[h_{\min} \le h(\vv{r}^{I}_k) \le
    h_{\max}\bigr],
  \label{eq:radio:altgate}
\end{equation}
the configured operating band of the instrument. The measurement is
computed and logged regardless of the flag, and draws are unconditional,
per the sensor-wide semantics of Chapter~\ref{ch:sensors-optical}.

\subsection{Acceptance statistics (exit criterion 6)}
\label{sec:radio:stats}

The gate scenario holds truth fixed, disables the correlated components
($\sigma_{\mathrm{GM}} = 0$ for the fix, $\sigma_b = 0$ for the
altimeter), and collects $M = 1000$ seeded samples. Per sample $k$ the
statistics are the exact normalized quadratic forms
\begin{equation}
  q^{(k)}_{r} = \sum_{a}
    \frac{(r_{\mathrm{meas},a} - r^{I}_a)^2_{(k)}}{\sigma_{r,a}^2}
    \sim \chi^2_3,
  \qquad
  q^{(k)}_{v} = \sum_{a}
    \frac{(v_{\mathrm{meas},a} - v^{I}_a)^2_{(k)}}{\sigma_{v,a}^2}
    \sim \chi^2_3,
  \qquad
  q^{(k)}_{h} =
    \frac{(h_{\mathrm{meas}} - h(\vv{r}^{I}))^2_{(k)}}{\sigma_{h}^2}
    \sim \chi^2_1,
  \label{eq:radio:chi2}
\end{equation}
each i.i.d.\ across samples by construction, and the gates are two-sided
on the ensemble means:
\begin{equation}
  \frac{\chi^2_{0.025}(nM)}{M} \le \frac{1}{M} \sum_{k=1}^{M} q^{(k)}
    \le \frac{\chi^2_{0.975}(nM)}{M},
  \qquad
  \begin{cases}
    [2.850,\; 3.154] & n = 3 \text{ (fix position, fix velocity)},\\[2pt]
    [0.914,\; 1.090] & n = 1 \text{ (altimeter)},
  \end{cases}
  \label{eq:radio:bounds}
\end{equation}
for $M = 1000$, quantiles evaluated by the Wilson--Hilferty form of
Chapter~\ref{ch:ekf}, equation~\eqref{eq:ekf:wh}. The quoted intervals are
the $M = 1000$ case, and the criterion's draw count is met by the C++ gate
\testid{sensors_navfix_altimeter_chi_square_over_1000_draws}. The pytest
driver recomputes the same statistic~\eqref{eq:radio:chi2} from the logged
\texttt{sensors.*} and \texttt{truth} channels, but over the $M = 300$
draws a \qty{60}{\second} mission at \qty{5}{\hertz} yields, against the
correspondingly wider bounds this same formula gives at that $M$ --- the
statistic is identical, the interval is not, and the two must not be
quoted interchangeably. The bias and
Gauss--Markov terms are pinned separately by deterministic tests (the
seeded draw values themselves), because a per-run bias makes the
per-sample statistics of equation~\eqref{eq:radio:chi2} dependent and
the mean gate invalid.

\section{Discretization and implementation}
\label{sec:radio:implementation}

Binding implementation notes for the Phase~6 modules:
\begin{enumerate}
  \item \textbf{Modules and traceability.} Both sensors are core-side
    \texttt{ISensor} modules on the named streams
    \texttt{sensors.navfix} and \texttt{sensors.alt} (D-9). Source
    comments echo \texttt{eq:radio:fix}, \texttt{eq:radio:gm},
    \texttt{eq:radio:alt}, and \texttt{eq:radio:chi2} verbatim.
  \item \textbf{Draw schedule, normative.} Nav fix, per sample and in
    order: position Gauss--Markov drive $(x,y,z)$ if enabled, velocity
    Gauss--Markov drive $(x,y,z)$ if enabled, position white $(x,y,z)$,
    velocity white $(x,y,z)$. Enabling a Gauss--Markov component also
    consumes its stationary-initialization draws at run start (position
    then velocity). Altimeter: turn-on bias (one draw at run start),
    then one white draw per sample. Within a fixed configuration the
    schedule is unconditional; enabling or disabling the optional
    components is a configuration change and lands in the resolved
    config hash (FR-15), which is the reproducibility anchor.
  \item \textbf{Geodetic reuse.} The altimeter calls the same core
    geodetic-altitude function the atmosphere and environment models
    use (Chapters~\ref{ch:harrispriester}, \ref{ch:environment}) --- no
    second conversion path; its accuracy budget
    (sub-millimetre in-domain) is inherited and already documented
    there.
  \item \textbf{Logged channels.} Measurements and flags land under
    \texttt{sensors.navfix.*} and \texttt{sensors.alt.*} with
    unit-suffixed names at the sensor rate.
\end{enumerate}

\section{Validation evidence}
\label{sec:radio:validation}

The module implementing this chapter is
\texttt{cpp/src/sensors/radio.cpp}.

\begin{longtable}{lp{8.2cm}}
  \caption{Validation evidence for the radio-class sensors: core (doctest,
    \texttt{cpp/tests/test\_sensors.cpp}).}
  \label{tab:radio:validation} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endfirsthead
  \multicolumn{2}{@{}l}{\itshape Table \thetable\ continued from the previous page.} \\
  \toprule
  Test ID & Acceptance basis \\
  \midrule
  \endhead
  \bottomrule
  \endlastfoot
    \testid{sensors_navfix_altimeter_chi_square_over_1000_draws} &
      Exit criterion~6: the three ensemble-mean
      statistics~\eqref{eq:radio:chi2} over $M = 1000$ seeded draws at
      fixed truth, with the correlated components disabled as
      Section~\ref{sec:radio:stats} requires, fall inside
      the~\eqref{eq:radio:bounds} intervals --- $[2.850, 3.154]$ for the
      position and velocity fixes and $[0.914, 1.090]$ for the
      altimeter. \\
    \testid{sensors_navfix_and_altimeter_error_terms} &
      Equations~\eqref{eq:radio:fix} and~\eqref{eq:radio:alt} value-for-value
      against independently replicated \texttt{sensors.navfix} and
      \texttt{sensors.altimeter} streams, which pins the normative
      per-sample draw schedules (position white then velocity white with
      the correlated components off; one turn-on bias draw at construction
      then one white draw per sample for the altimeter) as well as the
      arithmetic. The altimeter's turn-on bias is checked to be exactly the
      seeded draw, and the spherical-body altitude to be exactly
      $\norm{\vv{r}} - a$. \\
    \testid{sensors_navfix_gauss_markov_correlated_errors} &
      Equation~\eqref{eq:radio:gm} on the fix, value for value and then
      statistically. The white sigmas are zero, so the measured fix minus
      truth is exactly the correlated component, while the white draws are
      still consumed --- which is what makes the replayed schedule the
      normative one rather than a simplified one. Twenty-five samples are
      replayed against an independently constructed stream, pinning the
      construction-time stationary initialization (position then velocity)
      and the per-sample order (position drive, velocity drive, position
      white, velocity white). Over \num{20000} samples at
      $\Delta t / \tau = 0.025$ the sample variance sits in
      $[0.85, 1.15]\,\sigma^2$, which is the independent check of the drive
      relation $\sigma_w = \sigma\sqrt{1 - \phi^2}$ together with the
      stationary initialization, and the lag-one autocorrelation recovers
      $\phi = e^{-\Delta t / \tau}$ to \qty{2}{\percent}, which is the check
      that the configured $\tau$ reaches the recursion. A zero $\tau$ is
      confirmed to leave the component off with $\sigma$ alone
      configured. \\
    \testid{sensors_parsers_reject_an_unknown_parameter_name} &
      The allow-list in each radio parser: one misspelling of that sensor's
      own vocabulary is refused rather than ignored, after the same
      configuration without it is asserted to be accepted. \\
    \testid{sensors_factory_kind_vocabulary} &
      Both radio kinds construct from the factory at their configured rate
      and report their own \texttt{sensors.*} group names. \\
\end{longtable}

\begin{table}[htbp]
  \centering
  \caption{Validation evidence for the radio-class sensors: in-loop
    behavior and the blind reference re-gates (pytest,
    \texttt{tests/python/test\_gnc\_missions.py} and
    \texttt{test\_p6\_optical\_gates.py}).}
  \label{tab:radio:validation-py}
  \begin{tabular}{lp{8.2cm}}
    \toprule
    Test ID & Acceptance basis \\
    \midrule
    \testid{test_full_sensor_suite_reruns_bit_identical} &
      Exit criterion~1 bit identity, radio half. Two seeded runs of the
      reference mission with all six FR-23 sensor kinds declared --- the
      \texttt{sensors.navfix} and \texttt{sensors.altimeter} declarations
      among them, both carrying nonzero sigmas --- agree on the whole-file
      SHA-256 and then channel by channel under array bit equality, so both
      radio streams are pinned individually and not only in aggregate. \\
    \testid{test_altimeter_statistic_passes_the_reference_gate} &
      Equation~\eqref{eq:radio:alt} re-gated end to end through
      \texttt{tests/refs/sensor\_stats.py}, whose truth altitude is
      recomputed from the logged inertial position through the body-fixed
      rotation and the WGS84 ellipsoid. The fixture is inclined
      \qty{45}{\degree} deliberately: on an equatorial orbit the geodetic
      altitude degenerates to $\norm{\vv{r}} - a$ and the gate could not
      detect an implementation that never applied the ellipsoid. \\
    \testid{test_nav_fix_statistics_pass_the_reference_gate} &
      Equation~\eqref{eq:radio:fix} likewise, with the position and
      velocity components gated as separate $\chi^2_3$ statistics rather
      than one $\chi^2_6$, so a fix wired to the wrong truth row cannot be
      diluted by a healthy companion component. \\
    \bottomrule
  \end{tabular}
\end{table}

The altimeter's gating~\eqref{eq:radio:altgate} is exercised by the second
test above across the $h_{\max}$ boundary, with the white draw shown to be
consumed identically on both sides. The Gauss--Markov component of
equation~\eqref{eq:radio:gm} on the fix now has its own value-for-value
case, tabled above; until it was committed the whole model --- both
stationary initializations, both recursions, and \texttt{advance\_gm}
itself --- was entered by no test in either tier, while its three
parameters were documented mission configuration
(\texttt{docs/review/phase6\_coverage\_audit.md}, finding~C-3). Outstanding,
recorded rather than tabled: the altimeter's own
ellipsoidal $h = 0$ surface-point case is not yet committed --- the
altimeter-level case that is committed is the spherical one, asserting
$h = \norm{\vv{r}} - a$ exactly.

The ellipsoidal path is not thereby unpinned, and the previous sentence
should not be read as saying so. The conversion itself is gated by
\testid{ATM-HP-OFFNODE}, which synthesizes Earth-fixed points from exact
geodetic coordinates at four latitudes and recovers the altitude to
$< \qty{e-6}{\metre}$; and at sensor level
\testid{test_altimeter_statistic_passes_the_reference_gate} runs on a
\qty{45}{\degree}-inclined fixture chosen precisely so that the geodetic
and spherical altitudes differ by tens of metres against a
\qty{0.5}{\metre} sigma, so an implementation that never applied the
ellipsoid is rejected there. What is missing is the boundary case, not
the model.

\paragraph{Correction to assumption 4.} That assumption states that for a
spherical body ($f = 0$) the Bowring conversion of
equation~\eqref{eq:hp:geodetic} "degenerates exactly to
$h = \norm{\vv{r}} - a$." That is true of the mathematics but \emph{not}
of the implementation: \texttt{geodetic\_altitude()} validates
$1/f > 1$ and rejects zero flattening outright rather than returning the
limit. The altimeter therefore branches explicitly on a spherical central
body and evaluates $\norm{\vv{r}} - a$ in closed form, which additionally
needs no body-fixed rotation. The two branches agree in the limit and the
single-conversion-path intent of implementation note 3 is preserved for
every ellipsoidal body; the flat statement above should be read as
"degenerates to" in the mathematical sense only.

\section{References}
\label{sec:radio:references}

The fix-level abstraction of a navigation service and the additive
Gaussian error convention follow standard aided-navigation practice
(Titterton and Weston~\cite{titterton2004}); the Gauss--Markov machinery
is specified in Chapter~\ref{ch:sensors-imu} after
Maybeck~\cite{maybeck1979}; the geodetic conversion is
Bowring~\cite{bowring1976} as implemented in
Chapter~\ref{ch:harrispriester}. Full bibliographic details appear in
the bibliography.
